---
title: "Exercise_ADH"
output: html_document
date: '2022-09-05'
---

### Loading data, packages

```{r}
library(readr)
library(tibble)
library(fixest)
library(stringr)
library(dplyr)

library(lmtest)
library(sandwich)
```

```{r}
df = read_csv("adh_data.csv")
head(df)
```

### OLS regression

The core of the paper is looking at what happened to laborer's when theres an increase in us imports from china. 
Let's try and replicate part of Table 9 - namely the estimate from panel A column 2.

Their y variable is `relchg_avg_hhincwage_pc_pw`.

The important x variable is decadal trade between the us and china `d_tradeusch_pw`.

1. Run that simple regression
```{r}
    ols_1 = lm(relchg_avg_hhincwage_pc_pw ~ d_tradeusch_pw, data = df)
    summary(ols_1)
```

2. Now add heteroskedasticity robust standard  (HC1). Hint: Use the `sandwich` and `lmtest` packages
```{r}
    # ols_1 = lm(relchg_avg_hhincwage_pc_pw ~ d_tradeusch_pw, data = df)
    lmtest::coeftest(ols_1, sandwich::vcovHC(ols_1, "HC1"))
```

```{r}
ols_rob = feols(relchg_avg_hhincwage_pc_pw ~ d_tradeusch_pw, data = df, se = "HC1")
summary(ols_rob)
```

Now we will start to add extra x variables.

3. Start by adding `t2` - a dummy variable for whether observation is in the second decade. 
Fit again with HC1 robust standard errors.
```{r}
ols_2 = feols(relchg_avg_hhincwage_pc_pw ~ d_tradeusch_pw + t2,
                data = df, vcov = "HC1")
summary(ols_2)
```

```{r}
ols_2_hc3 = feols(relchg_avg_hhincwage_pc_pw ~ d_tradeusch_pw + t2,
                  data = df, vcov = function (x) sandwich::vcovHC(x, "HC3"))
summary(ols_2_hc3)
```

###  Clustering

Let us now use clustertered standard errors instead. ADH cluster by `statefip`.

1. Run the basic regression with clustering
```{r}
ols_clust = feols(relchg_avg_hhincwage_pc_pw ~ d_tradeusch_pw,
                  data = df, cluster = c("statefip"))
summary(ols_clust)
```
2. Add the following controls to your last regression:
    - `l_shind_manuf_cbp`
    - `l_sh_popedu_c`
    - `l_sh_popfborn`
    - `l_sh_empl_f`
    - `l_sh_routine33`
    - `l_task_outsource`
```{r}
ols_clust = feols(relchg_avg_hhincwage_pc_pw ~ d_tradeusch_pw + l_shind_manuf_cbp +
                    l_sh_popedu_c + l_sh_popfborn + l_sh_popfborn + l_sh_empl_f +
                    l_sh_routine33 + l_task_outsource,
                  data = df, cluster = c("statefip"))
summary(ols_clust)
```

3. Add region fixed effects to your regression.

    - First find all variables in the dataset that start with `reg_`
    - Add these to your last regression

```{r}
# reg_vars <- df %>% select(starts_with("reg_")) %>% names()
reg_vars <- df %>% colnames() %>% str_subset("reg_")

fe_str = str_c(reg_vars, collapse = " + ")
form_str = str_c(
    "relchg_avg_hhincwage_pc_pw ~ d_tradeusch_pw + l_shind_manuf_cbp",
    "l_sh_popedu_c + l_sh_popfborn + l_sh_popfborn + l_sh_empl_f",
    "l_sh_routine33 + l_task_outsource",
    fe_str,
    sep = " + "
)
form <- as.formula(form_str)

fe_model = feols(form, data = df, cluster = c("statefip"))
summary(fe_model)
```

```{r}
df = df %>% mutate(region = case_when(
  reg_midatl == 1 ~ 'midatl',
  reg_encen  == 1 ~ 'encen',
  reg_wncen  == 1 ~ 'wncen',
  reg_satl   == 1 ~ 'satl',
  reg_escen  == 1 ~ 'escen',
  reg_wscen  == 1 ~ 'wscen',
  reg_mount  == 1 ~ 'mount',
  reg_pacif  == 1 ~ 'pacif',
  TRUE ~ NA_character_
))

fe_model <- feols(
    relchg_avg_hhincwage_pc_pw ~ d_tradeusch_pw + l_shind_manuf_cbp +
        l_sh_popedu_c + l_sh_popfborn + l_sh_popfborn + l_sh_empl_f +
        l_sh_routine33 + l_task_outsource + i(region),
    data = df, cluster = c("statefip")
)
summary(fe_model)

fe_model <- feols(
    relchg_avg_hhincwage_pc_pw ~ d_tradeusch_pw + l_shind_manuf_cbp +
        l_sh_popedu_c + l_sh_popfborn + l_sh_popfborn + l_sh_empl_f +
        l_sh_routine33 + l_task_outsource | region,
    data = df, cluster = c("statefip")
)
summary(fe_model)
```

  
### Instrument Variables

1. Instrument `d_tradeusch_pw` with `d_tradeotch_pw_lag` in your last regression
```{r}
iv_reg = feols(
    relchg_avg_hhincwage_pc_pw ~ l_shind_manuf_cbp +
        l_sh_popedu_c + l_sh_popfborn + l_sh_popfborn + l_sh_empl_f +
        l_sh_routine33 + l_task_outsource | 0 | d_tradeusch_pw ~ d_tradeotch_pw_lag,
    data = df, cluster = c("statefip")
)
summary(iv_reg)
```

2. Weight your regression by `timepwt48`
```{r}
iv_reg_2 = feols(
    relchg_avg_hhincwage_pc_pw ~ l_shind_manuf_cbp +
        l_sh_popedu_c + l_sh_popfborn + l_sh_popfborn + l_sh_empl_f +
        l_sh_routine33 + l_task_outsource | 0 | d_tradeusch_pw ~ d_tradeotch_pw_lag,
    data = df, cluster = c("statefip"),
    weights = ~timepwt48
)
summary(iv_reg_2)
```

```{r}
iv_reg_2$iv_first_stage
```