---
title: "Exercise_ADH"
output: html_document
date: '2022-09-05'
---

---
title: "ADH_exercise"
output: html_document
date: '2022-09-05'
---

```{r}
library("readr")
library("tibble")
library("fixest")
library(stringr)
library(dplyr)
library(lmtest)
library(sandwich)
```

### OLS regression

The core of the paper is looking at what happened to laborer's when theres an increase in us imports from china. 
Let's try and replicate part of Table 9 - namely the estimate from panel A column 2.

```{r}
df = read_csv("adh_data.csv")
```

Their y variable is `relchg_avg_hhincwage_pc_pw`. 
The important x variable is decadal trade between the us and china `d_tradeusch_pw`.

1. Run that simple regression

```{r}
ols1 = lm(relchg_med_hhincwage_pc_pw ~ d_tradeusch_pw, data=df)
summary(ols1)
```

2. Now add heteroskedasticity robust standard  (HC1). Hint: Use the `sandwich` and `lmtest` packages

```{r}
ols_rob = feols(relchg_med_hhincwage_pc_pw ~ d_tradeusch_pw, se = 'hetero', data = df)
summary(ols_rob)

```


Now we will start to add extra x variables.

3. Start by adding `t2` - a dummy variable for whether observation is in the second decade. 
Fit again with HC1 robust standard errors.

```{r}
ols_t2 =  feols(relchg_med_hhincwage_pc_pw ~ d_tradeusch_pw + factor(t2), se = 'hetero', data = df)
summary(ols_t2)
```

###  Clustering

Let us now use clustertered standard errors instead. ADH cluster by `statefip`.

1. Run the basic regression with clustering

```{r}
ols_cluster = feols(relchg_med_hhincwage_pc_pw ~ d_tradeusch_pw, cluster = c('statefip'), data = df)
summary(ols_cluster)
```

2. Add the following controls to your last regression:
    - `l_shind_manuf_cbp`
    - `l_sh_popedu_c`
    - `l_sh_popfborn`
    - `l_sh_empl_f`
    - `l_sh_routine33`
    - `l_task_outsource`
    
```{r}
ols_cluster_control =  feols(relchg_med_hhincwage_pc_pw ~ d_tradeusch_pw + l_shind_manuf_cbp + l_sh_popedu_c + l_sh_popfborn + l_sh_empl_f + l_sh_routine33 + l_task_outsource,
                             cluster = c('statefip'), data = df)
summary(ols_cluster_control)
```
  

3. Add region fixed effects to your regression.
    - First find all variables in the dataset that start with `reg_`
    - Add these to your last regression

```{r}
stringv = str_split(names(df[ , grepl( "reg" , names( df ) ) ]), pattern = " ")

controls = str_c("relchg_med_hhincwage_pc_pw ~ d_tradeusch_pw + l_shind_manuf_cbp + l_sh_popedu_c + l_sh_popfborn + l_sh_empl_f + l_sh_routine33 + l_task_outsource ",
                 str_c(stringv, collapse = " + "), sep = ' + ')


ols_cluster_fe =  feols( as.formula(controls), cluster = c('statefip'), data = df)
summary(ols_cluster_fe)
```

We can also do this with a function in case we have to do it many times:
```{r}

# Function for formulas ---------------------------------------------------
# sep	: String to insert between input vectors.
# collapse	: Optional string used to combine input vectors into single string.
create_formulas <- function(dep_var, treat, expl_var) {
  controls <- str_c(expl_var, collapse  = '+')
  fmla <- str_c(dep_var, treat, sep = '~') %>% 
          str_c(controls, sep = ' + ')
  return(as.formula(fmla))
}

y <- 'relchg_med_hhincwage_pc_pw'
d <- 'd_tradeusch_pw'
X <- c('l_shind_manuf_cbp' , 'l_sh_popedu_c' , 'l_sh_popfborn' , 'l_sh_empl_f' , 'l_sh_routine33' , 'l_task_outsource')
  
X_new <- df %>% colnames() %>% .[stringr::str_detect(.,'reg_')] %>%  c(X)

create_formulas(y,d, X_new)

lm_formula = feols(create_formulas(y,d, X_new), data=df)
summary(lm_formula)
```

Create region dummy (string) 

```{r}
df = df %>% mutate(region = case_when(
  reg_midatl == 1 ~ 'midatl',
  reg_encen  == 1 ~ 'encen',
  reg_wncen  == 1 ~ 'wncen',
  reg_satl   == 1 ~ 'satl',
  reg_escen  == 1 ~ 'escen',
  reg_wscen  == 1 ~ 'wscen',
  reg_mount  == 1 ~ 'mount',
  reg_pacif  == 1 ~ 'pacif',
  TRUE ~ NA_character_
))

#The two regressions below are the same:

lm_1 = feols(relchg_med_hhincwage_pc_pw ~  d_tradeusch_pw + l_shind_manuf_cbp + l_sh_popedu_c + 
             l_sh_popfborn + l_sh_empl_f + l_sh_routine33 + l_task_outsource | region,  vcov = 'iid', data=df)

summary(lm_1)

lm_2 = feols(relchg_med_hhincwage_pc_pw ~  d_tradeusch_pw + l_shind_manuf_cbp + l_sh_popedu_c + 
             l_sh_popfborn + l_sh_empl_f + l_sh_routine33 + l_task_outsource + i(region), vcov = 'iid', data=df)
summary(lm_1)

```

### Instrument Variables

1. Instrument `d_tradeusch_pw` with `d_tradeotch_pw_lag` in your last regression

```{r}

IV = feols(relchg_med_hhincwage_pc_pw ~  l_shind_manuf_cbp + l_sh_popedu_c + 
             l_sh_popfborn + l_sh_empl_f + l_sh_routine33 + l_task_outsource | 0 
            | d_tradeusch_pw ~ d_tradeotch_pw_lag ,
             cluster = c('statefip'), data = df)
summary(IV)
```

2. Weight your regression by `timepwt48`

```{r}
IV_weighted = feols(relchg_med_hhincwage_pc_pw ~  l_shind_manuf_cbp + l_sh_popedu_c + 
                      l_sh_popfborn + l_sh_empl_f + l_sh_routine33 + l_task_outsource | 0 
                    | d_tradeusch_pw ~ d_tradeotch_pw_lag ,
                    cluster = c('statefip'), weights = ~timepwt48 ,data = df)
summary(IV_weighted)
```
