---
title: "Factors_and_Strings"
output: html_document
date: '2022-08-28'
---

This file is greatly based on the material prepared by Julian Langer and Lachlan Deer for the programming course for previous years

## Working with strings

```{r}
library(stringr)
library(tibble)
library(dplyr)
library(ggplot2)
```

## String combine with `str_c`
```{r}
our_string = c("With", "great", "power", "comes", "great", "responsibility")
str_length(our_string)
length(our_string) #note the difference between length and str_length
```
Pasting together strings

```{r}
str_c("Hello", "World", sep = " ")
```
```{r}
str_c(our_string,  collapse = " ")
```
## Substrings  with `str_sub`: str_sub(string, start = 1L, end = -1L)
```{r}
str_sub(our_string, 2, 4)
str_sub(our_string, -1, -1)
str_sub(our_string, -2, -1)
```
## Capitalize or not
```{r}
str_to_lower(our_string)
str_to_upper(our_string)
str_to_title(our_string)
# specify locale to be sure that behavior is consistent with locale option
str_to_title(our_string, locale = "en")
```
## String sort
```{r}
str_sort(our_string)
```
## Handling NAs
```{r}
another_string = c(NA, "Hello", "World")
str_c(another_string, collapse = " ")
str_c(str_replace_na(another_string, replacement = ""), collapse = " ")
str_c(na.omit(another_string), collapse = " ")
```
## Finding an expression in a string

```{r}
str_detect(our_string, "a")
str_detect(our_string, "a|W")
str_detect(our_string, "[aeiou]")
str_detect(c('8','1'), "[0-5]")
str_detect(our_string, "s$") # s at end of string
str_detect(our_string, "^G") # G at beginning of string
```
A bit more tricky: Find expressions with an o followed by an s, with any number of characters in the middle (a sequence of at least one character that isn’t a space: [^ ]+)

```{r}
our_string = c("oas", "os", "oa", "ot", "olkjdnsa", "so")
str_detect(our_string, "o[^ ]+s|os") #  o followed by s
```

## Return whole strings containing a certain expression
```{r}
str_subset(our_string, "a")
```

## Replacing an expression in a string

```{r}
str_replace(our_string, 'a', 'L')
```

## Working with factors

In R, factors are used to work with categorical variables, variables that have a fixed and known set of possible values. They are also useful when you want to display character vectors in a non-alphabetical order

Factors are the data objects which are used to categorize the data and store it as levels. They can store both strings and integers


## Creating Factors

```{r}
my_dataframe = tibble(
  age = c(12, 17, 11, 8),
  gender = c("Male", "Male", "Female", "Female"),
  review = c("C", "B", "B", "A")
)
```

```{r}
my_dataframe = my_dataframe %>%
  mutate(gender = factor(gender, levels = c("Male", "Female"))) %>% 
  mutate(review = factor(review, levels = c("C", "B", "A"), ordered = TRUE))
head(my_dataframe)
```
So, we created an ordered as well as an unordered factor. If you want to see the values for the factor levels, use `levels`. 
```{r}
levels(my_dataframe$gender)

levels(my_dataframe$review)
```
Why would we want an ordered factor? One reason is that we can filter using '>' or '<' based on ordered factors
Another reason is that they can be very useful when we want to make plots, and we want the variables in the x axis to have a specific order

## Some other useful functions from forcats
```{r}
library(forcats)
```

### Modify factor levels 
```{r}
my_dataframe = my_dataframe %>%
  mutate(review = fct_recode(review,
                             "Very Good" = "A",
                             "Good"      = "B", 
                             "Bad"       = "C"))
head(my_dataframe)
```

No bad people.

```{r}
my_dataframe = my_dataframe %>%
  mutate(review = fct_recode(review,
                             "Good" = "Bad"))
head(my_dataframe)
```

# Ordered factors are very useful for plotting

Now we use on forcats::gss_cat. It’s a sample of data from the General Social Survey, which is a long-running US survey conducted by the independent research organization NORC at the University of Chicago

```{r}
gss_cat

```
```{r}
relig_summary <- gss_cat %>%
  group_by(relig) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )
#> `summarise()` ungrouping output (override with `.groups` argument)
levels(relig_summary$relig)

ggplot(relig_summary, aes(tvhours, relig)) + geom_point()
```

```{r}
ggplot(relig_summary, aes(tvhours, fct_reorder(relig, tvhours))) +
  geom_point()
```
```{r}
gss_cat %>% filter(relig == 'Protestant')
```
factors are also useful to include categorical variables in a regression

```{r}
lm(tvhours ~ age, data = gss_cat)
lm(tvhours ~ as.factor(age), data = gss_cat)
```

