---
title: "ReadR_Parsing"
output: html_document
date: '2022-08-15'
---

This file is greatly based on the material prepared by Julian Langer and Lachlan Deer for the programming course for previous years

## Reading data


In this section we will learn how to read in and export data in different formats 

### csv files 
One of the most common files you'll have to import with R are csv (comma-separated-values) files.

Base R allows you to read csv files with the "read.csv" command. However, we will prefer another command from the "readr" library. 

"read.csv" opens the data as a data.frame, the package "readr" opens data as a tibble, which we prefer as discussed before. Also, tibbles load faster than data.frames

```{r, message = F}
library(readr)
library(tibble)
library(dplyr)
getwd() # let's check the working directory
df_adh <- read_csv("../Datasets/adh_data.csv")
```

### Reading delimited data 
More generally, "readr" has functions to read in other types of delimited data

* `read_csv()`: comma delimited
* `read_csv2()`: semicolon delimited
* `read_tsv()`: tab delimited
* `read_delim()`: any delimiter

For instance:

```{r, message = F}
df_adh <- read_delim("../Datasets/adh_data.csv", delim = ",")
```

## Parsing data correctly with the `parse` functions

Sometimes columns are not correctly interpreted when they are read in. For these vectors, we can parse them differently using the `parse` functions.  These functions take a character vector and return a more specialised vector like a logical, integer, or date. For example, assume we want to retrieve the numbers of the following column

```{r, error=TRUE}
my_column = c("100$", "20%", "Something with a 100")
parse_number(my_column)
```
There are a bunch of functions, each for a different kind of data type:  

* `parse_logical`
* `parse_number` (`parse_double`, `parse_integer`)
* `parse_character`
* `parse_datetime` (`parse_date`, `parse_datetime`)

Here are some more examples


```{r}
parse_logical(c("true", "false", "t", "T", "f"))
```
```{r}
parse_number(c("$1,234.5", "$12.45"))
```
What if we have a file with the comma and the dot inverted?

```{r}
parse_number(c("$1.234,5", "$12,45"), locale = locale(decimal_mark = ","))
```

```{r}
# Used in America
parse_number("$123,456,789")
#> [1] 123456789

# Used in many parts of Europe
parse_number("123.456.789", locale = locale(grouping_mark = "."))
#> [1] 123456789

# Used in Switzerland
parse_number("123'456'789", locale = locale(grouping_mark = "'"))
#> [1] 123456789
```

**Parsing characters**

You would think parsing characters is pretty easy. There can be difficulties though because not everybody uses the same character encoding. To learn more about this topic, take a look at this website: http://www.w3.org/International/articles/definitions-characters/. We only need to know that different encodings exist and they can lead to problems. R usually assumes to 'UTF-8' encoding (and you should use it too!). See what happens if we read in characters with `Latin-1` encoding: 

```{r}
x1 = "El Ni\xf1o was particularly bad this year"
parse_character(x1)
```
Well, that does not look nice. But luckily enough, we can use the `locale` function to tell `readr` that the string is encoded with `Latin-1`. 
```{r}
parse_character(x1, locale = locale(encoding = "Latin1"))
```
Now, this time the parsing is correct! You can also use `readr` to try to guess the encoding with the `guess_encoding` function. Look up its help file if you want to know more.  

** Parsing dates and times **
Sometimes it is important for us to read certain data as date or time. We can also to this by parsing:

```{r}
parse_date("2016-09-08") # default option is Y-m-d
parse_date("2016/02/29") + 1 # It can distinguish leap years!
```

```{r}
parse_datetime("2016-09-08T0708")
parse_datetime("20160908T0708")

```
date_format specified by the locale(). The default value is %AD which uses an automatic date parser that recognises dates of the format Y-m-d or Y/m/d.

```{r}
parse_date("27/05/1987", "%d/%m/%Y")
```
Now let's assume we want to extract the month, we can do:

```{r}
format(parse_date("27/05/1987", "%d/%m/%Y"), "%m")
```
```{r}
library(hms)
parse_time("01:10 am") 
```

## Parsing and reading at the same time

Each `parse` function has a corresponding `col` function. This allows you to use the `parse` function to find out how to correctly parse a column and then specify the correct parsing right at the beginning of the data processing using the corresponding `col` function. I usually read in data in three steps.

1. First, I read in all columns as character vectors. This allows me to browse the data and determine the correct parsing. To read in every column as a character vector, you can use the `.default` argument in the `col_types` function. 

```{r}
challenge1 = read_csv(readr_example("challenge.csv"),  
                      # there's an example dataset in the readr package called challenge.csv
                     col_types = cols(
                       .default = col_character()
                     ))
```

2. I can try out different parsers using the `parse` functions. (Note that you can use the `parse` functions from the `readr` package together with other packages such as `readxl`). In this case, browsing and parsing will lead you to conclude that the correct parsers are `parse_double` and `parse_date`, respectively. 

3. Finally, we specify the correct parsers directly at the beginning of the data processing stage using the `col` functions that correspond to the `parse` functions. 
```{r}
challenge2 = read_csv(readr_example("challenge.csv"),
                      col_types = cols(
                        x = col_double(),
                        y = col_date()
                      ))
head(challenge2)
```

## Writing `csv` files
You can use the `readr` package to write `csv` files as well. In this case, we want to save our cleaned up dataframe as a `csv` file in a `dataframes` sub-folder. For this, we first check whether the sub-folder already exists. If it does not, we create it. 
```{r}
if (!file.exists("../new_dataframes")) {
  dir.create("../new_dataframes")
}
```
Then, we write the `csv` file using the `write_csv` function. 
```{r}
#write_csv(challenge2, "../dataframes/challenge2.csv")
```
Note that a `csv` file does not store the information about the correct parsing of the data columns.

We have to specify the correct parsing again! If you only work with `R` and the dataframe is not too big, you can store the dataframe as an `RDS` file instead.

## Write and read `RDS` files with `readr`
There is not much to say here apart from the fact that the `RDS` file 'remembers' the correct parsing. 
```{r}
#write_rds(challenge2, "./dataframes/challenge2.rds")
#challenge_rds = read_rds("./dataframes/challenge2.rds")
#head(challenge_rds)
```

## Reading in Excel sheets and Stata data with `readxl` and `haven` 
Many datasets are stored in Excel sheets. You can read them in using the `readxl` package. A similar package, `haven`, exists for Stata files. I will not use them here, but can look at the helpfiles for the commands `read_excel` and `read_dta`. 
```{r}
library(readxl)
help(read_excel)
```

```{r, message = FALSE}
library(haven)
help(read_dta)
```

If you want to try them out, you can use the data files from the 'Baby-Wooldridge' here: http://www.cengage.com/aise/economics/wooldridge_3e_datasets/. 






